---
title: "Behind Bars and Beyond: Investigating Socioeconomic Drivers of Incarceration Rates in the United States"
subtitle: "A comparative analysis of global trends and America's mass incarceration crisis."
author: 
  - Abbass Sleiman
thanks: "Code and data are available at: [https://github.com/AbbassSleiman/US_Incarceration](https://github.com/AbbassSleiman/US_Incarceration). Shiny web app can be accessed at [https://wklzec-abbass-sleiman.shinyapps.io/incarceration_rate/](https://wklzec-abbass-sleiman.shinyapps.io/incarceration_rate/)"
date: today
date-format: long
format: 
  pdf: 
    include-in-header: 
      text: |
        \usepackage{lscape}
        \newcommand{\blandscape}{\begin{landscape}}
        \newcommand{\elandscape}{\end{landscape}}
    documentclass: article
    geometry: margin = 1.2in
    abstract: "Over the last 2 decades, the United States has found itself struggling with an issue of mass incarceration, leading the world in number of individuals incarcerated. This paper investigates the possible causes of this by conducting a comparative analysis of global trends in education, unemployment, violent crime, and poverty rates, and their respective impacts on incarceration. Utilizing data from the World Population Review, and employing a Bayesian multiple linear regression model, this paper finds evidence that poverty and unemployment negatively correlate with incarceration rates, whereas violent crime rates positively correlate. The paper also identifies nuances in the relationship between educational attainment and incarceration rates. These insights may offer valuable considerations for policy-making aimed at addressing mass incarceration in the United States."
    number-sections: true
output: 
  bookdown::pdf_document2
toc: true
bibliography: references.bib
---

```{r}
#| include: false
#| warning: false
#| message: false

# Package loading
library(tidyverse)
library(here)
library(ggplot2)
library(knitr)
library(modelsummary)
library(rstanarm)
library(kableExtra)
library(dplyr)
library(janitor)
library(shiny)

# Read data
raw_incarceration_data <- read_csv(here("data/raw_data/incarceration-rates-by-country-2024.csv"))
raw_poverty_data <- read_csv(here("data/raw_data/poverty-rate-by-country-2024.csv"))
raw_violent_crime_data <- read_csv(here("data/raw_data/violent-crime-rates-by-country-2024.csv"))
raw_unemployment_data <- read_csv(here("data/raw_data/unemployment-by-country-2024.csv"))
raw_education_data <- read_csv(here("data/raw_data/education-rankings-by-country-2024.csv"))

cleaned_data <- read_csv(here("data/analysis_data/incarceration_data.csv"))

# Read model
analysis_model <- readRDS(file = here("models/incarceration_model.rds"))

```


# Introduction

In recent decades, the United States has grappled with a consistent rise in its prison population, becoming infamous for having one of the highest incarceration rates in the world. Leading the world with nearly 2 million individuals in prisons and jails in 2022, the US is faced with what, at first glance, appears to be a "uniquely American problem" [@citeUSincarceration]. This phenomenon has sparked widespread concern and debate over how best to address this issue. Central to these discussions is the question of what underlying factors have contributed to this circumstance, in turn necessitating an exploration of the socioeconomic drivers of incarceration rates in the United States.

Incarceration refers to the state of being confined in prison as punishment for criminal offences [@citeIncarcerationRate], and in examining the United States' prison dilemma, it becomes clear that the crux of the problem is not easily reducible to simple explanations. Instead, it is a complex issue influenced by a multitude of factors including cultural dynamics and socioeconomic conditions. Thus, in an attempt to unravel the intricacies of these factors, this paper seeks to conduct a comparative analysis of global trends in education, unemployment, violent crime, and poverty rates as independent factors influencing a nation's incarceration. However, recognizing that the total number of incarcerations is heavily dependent on population size, this paper focuses on nations' incarceration rates as the estimand, defined to be the number of incarcerated individuals per 100,000 in a given population [@citeIncarcerationRate], which serves as a better metric for measuring the extent of imprisonment within a society. By exploring the impact of these factors on incarceration rates, we aim to better understand the primary forces behind the United States' high incarceration rate and overall number of individuals incarcerated. This understanding will help inform policy-making decisions aimed at addressing the issue effectively.

This paper utilizes various datasets from the World Population Review, in particular it utilizes their datasets on "Incarceration Rates by Country 2024", "Poverty Rate by Country 2024", "Education Rankings by Country 2024", "Unemployment by Country 2024", and "Violent Crime Rates by Country 2024". By conducting statistical analyses with the aforementioned data, this paper finds evidence to suggest that poverty and unemployment negatively correlate with incarceration rates, and violent crime rates correlate positively. Education, on the other hand, appears to have a more nuanced relationship with incarceration rates as the relationship does not appear to be consistent across all levels of education.  

The remainder of this paper is structured as follows. @sec-data discusses the raw data, cleaning process, variables of interest, and offers visual representations of the data through tabular and graphical representations of the observations. @sec-model introduces and justifies the Bayesian linear regression model used in the analysis of the data in predicting the relationship between the aforementioned independent variables and a nation's incarceration rate. @sec-results deals with analyzing the trends and correlations showcased by the model in more detail. @sec-discussion discusses the real-world implications of the results uncovered in the prior sections, and finally @sec-limitations discusses the limitations and weaknesses of the analysis conducted, as well as the next steps that could be taken to improve the overall reliability of the paper. Additional information regarding the model details can be found in the Appendix under @sec-model-details.

# Data {#sec-data}

This paper utilizes 5 datasets, all provided by the World Population Review and each providing information on one of the 5 factors involved in this study. The incarceration rate dataset [@citeIncarcerationRate], @sec-incarceration-data, features the most recently available information on incarcerations and incarceration rates of 219 countries and territories as of 2024. The poverty rate dataset [@citePovertyRate], @sec-poverty-data, features the data on the poverty rates of 164 nations. The violent crime rate dataset [@citeViolentCrimeRate], @sec-violent-crime-data, features data on the number of various types of violent crimes in each of 200 nations and territories per 100k members of the population. The unemployment rate dataset [@citeUnemploymentRate], @sec-unemployment-data, features data on the unemployment rate in each of 226 countries and territories. Finally, the education ranking dataset [@citeEducationRanking], @sec-education-data, contains data on the literacy rate and education rank of each of 207 countries and territories, accurate as of 2024.

All data analysis was done through R [@citeR] with the aid of the following packages: `tidyverse` [@citetidyverse], `here` [@citehere], `ggplot2` [@citeggplot2], `knitr` [@citeknitr], `modelsummary` [@citemodelsummary], `rstanarm` [@citerstanarm], `kableExtra` [@citekableExtra], `dplyr` [@citedplyr], and `janitor` [@citejanitor]. Moreover, the shiny web app associated with this paper was made with the aid of the `shiny` package [@citeshiny].

## Incarceration Rate Dataset {#sec-incarceration-data}

The raw incarceration rate dataset contains information on the incarceration rate (per 100k population), total incarcerated, percentage of incarcerated that are female, and percentage of incarcerated that are male for 219 countries and territories. A snippet of the top 10 countries ordered by incarceration rate can be seen in @tbl-incarceration-data, showcasing their rates of incarceration and total number of individuals incarcerated. It is important to note that, whilst the data provided is the most recent available for each nation, the data for each nation need not be accurate as of 2024 as varying nations may have data from different years.

```{r}
#| label: tbl-incarceration-data
#| tbl-cap: Top 10 countries by incarceration rate
#| echo: false
#| tbl-pos: H

raw_incarceration_data |>
  arrange(desc(incarcerationRatesByCountry_ratePer100k)) |>
  head(10) |>
  select(country, incarcerationRatesByCountry_ratePer100k, incarcerationRatesByCountry_totIncarcerated) |>
  kable(
    col.names = c("Country", "Incarceration Rate per 100k Population", "Total Incarcerated"), 
    booktabs = TRUE
  )
```

## Poverty Rate Dataset {#sec-poverty-data}

The poverty rate dataset contains information on the poverty rates in 164 nations. In particular, it provides information on the percentage of the population living on less than 10 USD per day, less than 3.65 USD per day, the poverty rate, and the year the data was collected. Here, poverty rate is measured as the percentage of a given population living below that particular nation's respective poverty line which is estimated based on population-weighted subgroup estimates from household surveys [@citePovertyRate]. This is an important clarification to make as it ensures that a nation's poverty rate is relative to the living costs in that particular nation, in turn reducing biases that could arise as a result of the vastly differing costs of living across various nations. This ensures that, in the analysis that shall follow, we are able to more accurately take into account income inequality relative to each particular nation. The top 5 countries ordered by poverty rate are showcased in @tbl-poverty-data, showcasing information on the poverty rate and year of data collection for each of the 5 countries.

```{r}
#| label: tbl-poverty-data
#| tbl-cap: Top 5 countries by poverty rate
#| echo: false
#| tbl-pos: H

raw_poverty_data |>
  arrange(desc(PovertyRateValue)) |>
  head(5) |>
  select(country, PovertyRateValue, PovertyRateValueDataYear) |>
  kable(
    col.names = c("Country", "Poverty Rate", "Year Data was Collected"), 
    booktabs = TRUE
  )
```

## Violent Crime Rate Dataset {#sec-violent-crime-data}

The violent crime rate dataset contains violent crime statistics for 200 nations and territories across the globe. In particular, it contains data on the homicide rate, serious assault rate, sexual violence rate, kidnapping rate, and robbery rate for each of the nations, each of which is measured to be the number of incidents per 100k population. It also features information on the year each of the data points was collected. As with the other datasets, the year of data collection for each nation varies by nation, and thus some information is more recent than others. 

An important note to make regarding this information is that the definitions for each of these crimes can vary significantly from nation to nation. What one nation may deem to be a serious offence can be nothing more than a misdemeanor in another. As a result, though the data provided is based on each nation's own definition of a serious offence, the fact that we are still provided with a fairly well-rounded subset of violent crimes ranging from theft to murder will enable us to better understand what the overall rate of violent crimes per nation is comprised of. One should note that some countries have missing values for certain crime rates, indicated by an "NA". @tbl-violent-crime-data showcases a sample of 5 countries including information regarding the aforementioned crime rates for each.

```{r}
#| label: tbl-violent-crime-data
#| tbl-cap: Sample of 5 countries and their respective rates of various violent crimes (per 100k population)
#| echo: false
#| tbl-pos: H

raw_violent_crime_data |>
  head(5) |>
  select(country, violentCrimeRatesByCountry_homRate, violentCrimeRatesByCountry_assault, violentCrimeRatesByCountry_sexViolence, violentCrimeRatesByCountry_kidnap, violentCrimeRatesByCountry_robbery) |>
  kable(
    col.names = c("Country", "Homicide Rate", "Serious Assault Rate", "Sexual Violence Rate", "Kidnapping Rate", "Robbery Rate"), booktabs = TRUE
  ) |>
  column_spec(1, width = "1.8cm") |>
  column_spec(2, width = "2.2cm") |>  # Adjust the width of the first column
  column_spec(3, width = "2.2cm") |>  # Adjust the width of the first column
  column_spec(4, width = "2.2cm") |>  # Adjust the width of the first column
  column_spec(5, width = "2.2cm") |>  # Adjust the width of the first column
  column_spec(6, width = "2.2cm")     # Adjust the width of the second column
```

## Unemployment Rate Dataset {#sec-unemployment-data}

The unemployment rate dataset provides data on the unemployment rates of 226 countries and territories across the globe according to 3 various sources: the World Bank, the International Labour Organization, and the CIA, as well as the respective years that each of these organizations collected the data. Here, unemployment is defined as the percentage of unemployed workers in the labour force, which includes those who do not currently work but are able to do so [@citeUnemploymentRate]. It is important to note that, as with the other datasets, information on the nations' unemployment rates is not necessarily accurate as of 2024 and the date of collection varies from country to country. As the World Bank has information on every one of the countries and territories, this paper will utilize only data from that one source regarding unemployment rates. @tbl-unemployment-data showcases the top 5 countries by highest unemployment rates, according to the World Bank, as well as the year the data was collected.

```{r}
#| label: tbl-unemployment-data
#| tbl-cap: Top 5 countries by highest unemployment rate (as per the World Bank)
#| echo: false
#| tbl-pos: H

raw_unemployment_data |>
  arrange(desc(UnemploymentRateWorldBank)) |>
  head(5) |>
  select(country, UnemploymentRateWorldBank,UnemploymentRateWorldBankDataYear) |>
  kable(
    col.names = c("Country", "Unemployment Rate", "Year Data was Collected"), 
    booktabs = TRUE
  )
```

## Education Ranking Dataset {#sec-education-data}

The education ranking dataset provides information on the literacy rate and education rank of 207 countries and territories. The dataset contains ranks assigned by various organizations, however the most recent ranking data is by the nonprofit World Top 20, who determine a nation's rank according to a number of factors including early childhood enrollment rates, high school graduation rates, primary test scores, and various other factors [@citeEducationCriteria]. The top 5 countries according to the World Top 20 in 2024 are showcased in @tbl-education-data.

```{r}
#| label: tbl-education-data
#| tbl-cap: Top 5 countries by education (World Top 20 (2024))
#| echo: false
#| tbl-pos: H

raw_education_data |>
  arrange(educationRankingsByCountry_WT20rank2024) |>
  head(5) |>
  select(country, educationRankingsByCountry_WT20rank2024) |>
  kable(
    col.names = c("Country", "Education Rank"), booktabs = TRUE
  )
```

## Cleaned Dataset {#sec-cleaned-data}

The finalized cleaned dataset used in the analysis that will follow was created by merging the aforementioned 5 datasets into one. In particular, only countries that were present in all 5 datasets are present in the cleaned dataset, bringing the total number of nations included in this study to 134. Further, the cleaned analysis dataset was filtered to only contain the following information from the 5 datasets: a country's name, its incarceration rate (per 100k population), its poverty rate, its unemployment rate, its violent crime rate, and its educational ranking category. Note that a nation's violent crime rate and educational ranking category were not explicit data that were provided as part of the 5 raw datasets. Instead, these were manually constructed and appended to the final dataset during the cleaning process. 

In this study, a nation's violent crime rate is defined to be the sum of all the crime rates (per 100k population) available through the violent crime dataset in order to have one final value that represents the extent of violent crime in a nation. It is important to note, however, that various nations had missing values for some of the violent crime statistics, and as a result, these values were simply treated as 0 for the sake of the summation. As for the educational ranking category, as opposed to treating educational ranking as a continuous variable in the analysis that will follow, we instead decided to create 10 categories representing the 10 deciles of educational ranks present in the final dataset, based on the ranks of the 134 countries present, essentially splitting the countries into 10 equally sized categories by educational ranking. This was done in order to have a categorical variable whose coefficient will be more easily interpreted in the model present in @sec-model, as interpreting a 1 unit increase in educational ranking can be quite arbitrary, given that moving up or down one rank educationally may be more subjective. @tbl-cleaned-data showcases the top 5 countries by incarceration rate in the cleaned dataset.

```{r}
#| label: tbl-cleaned-data
#| tbl-cap: Top 5 countries by incarceration rate in the finalized cleaned dataset
#| echo: false
#| tbl-pos: H

cleaned_data |>
  arrange(desc(incarceration_rate)) |>
  head(5) |>
  kable(
        col.names = c("Country", "Incarceration Rate", "Poverty Rate", "Unemployment Rate", "Educational Ranking Category", "Violent Crime Rate"), booktabs = TRUE
  ) |>
  column_spec(1, width = "2.2cm") |>
  column_spec(2, width = "2.2cm") |>  # Adjust the width of the first column
  column_spec(3, width = "2.2cm") |>  # Adjust the width of the first column
  column_spec(4, width = "2.2cm") |>  # Adjust the width of the first column
  column_spec(5, width = "2.2cm") |>  # Adjust the width of the first column
  column_spec(6, width = "2.2cm")
```

To gain a better understanding of the distribution of incarceration rates as a function of the independent variables, the figures below each showcase all observations, placing incarceration rates on the y-axis and independent variable on the x-axis. Moreover, we have chosen to highlight 5 particular nations in each of the figures, representing nations across the entire range of incarceration rates present in the dataset. In particular, we highlight El Salvador who has the highest incarceration rate of 1086, the United States with the 3rd highest incarceration rate of 531, Mexico with the 51st highest incarceration rate of 174, Iceland with the 5th lowest incarceration rate of 36, and Guinea-Bissau with the lowest incarceration rate of 31. Do note that these rankings are relative to the finalized dataset as opposed to the entire list of countries and territories found in the raw incarceration rate dataset. 

@fig-variable-incarceration showcases all observations by poverty rate, unemployment rate, and violent crime rates, highlighting the aforementioned countries as well. On the other hand, @fig-education-incarceration showcases the mean incarceration rate for each of the 10 educational ranking categories. Note that Iceland is in the "1-19" category, the United States is in the "20-34" category, Mexico is in the "35-52" category, El Salvador is in the "111-125" category, and Guinea-Bissau is in the "176-198" category.

```{r}
#| label: fig-variable-incarceration
#| fig-cap: Scatter plot of incarceration rates against each of the independent variables of interest.
#| fig-subcap: ["Poverty Rate", "Unemployment Rate", "Violent Crime Rate"]
#| echo: false
#| layout-ncol: 2
#| fig-pos: H

ggplot(cleaned_data, aes(x = poverty_rate, y = incarceration_rate)) +
  geom_point(color = "darkgrey") +
  geom_point(data = cleaned_data |> filter(country %in% c("United States", "El Salvador", "Guinea-Bissau", "Iceland", "Mexico")),
             aes(color = country), size = 1.5) +  # Highlight specific countries
  theme_classic() +
  labs(x = "Poverty Rate", y = "Incarceration Rate")

ggplot(cleaned_data, aes(x = unemployment_rate, y = incarceration_rate)) +
  geom_point(color = "darkgrey") +
  geom_point(data = cleaned_data |> filter(country %in% c("United States", "El Salvador", "Guinea-Bissau", "Iceland", "Mexico")),
             aes(color = country), size = 1.5) +  # Highlight specific countries
  theme_classic() +
  labs(x = "Unemployment Rate", y = "Incarceration Rate")

ggplot(cleaned_data, aes(x = violent_crime_rate, y = incarceration_rate)) +
  geom_point(color = "darkgrey") +
  geom_point(data = cleaned_data |> filter(country %in% c("United States", "El Salvador", "Guinea-Bissau", "Iceland", "Mexico")),
             aes(color = country), size = 1.5) +  # Highlight specific countries
  theme_classic() +
  labs(x = "Violent Crime Rate", y = "Incarceration Rate")
```

```{r}
#| label: fig-education-incarceration
#| fig-cap: Bar chart showcasing mean incarceration rate for each educational ranking category
#| echo: false
#| fig-pos: H

# Define the levels in the desired order
education_levels <- c("1-19", "20-34", "35-52", "53-71", "72-92", 
                      "93-110", "111-125", "126-147", "148-175", "176-198")

# Convert education_category to a factor with specified levels
cleaned_data$education_category <- factor(cleaned_data$education_category, levels = education_levels)

# Recalculate the mean incarceration rate for each educational category
mean_incarceration <- cleaned_data |>
  group_by(education_category) |>
  summarize(mean_incarceration_rate = mean(incarceration_rate, na.rm = TRUE))

# Create the bar chart
ggplot(mean_incarceration, aes(x = education_category, y = mean_incarceration_rate)) +
  geom_bar(stat = "identity", fill = "skyblue") +
  theme_classic() +
  labs(x = "Educational Ranking Category", y = "Mean Incarceration Rate") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))  # Rotate x-axis labels for better readability

```

To gain a better understanding of the underlying data, @tbl-summary-stats provides information on the overall mean, median, and standard deviation of each of the variables present in the cleaned dataset.

```{r}
#| label: tbl-summary-stats
#| tbl-cap: Mean, median, and standard deviation for each variable
#| echo: false
#| tbl-pos: H

# Calculate summary statistics
mean_incarceration <- round(mean(cleaned_data$incarceration_rate, na.rm = TRUE), 2)
sd_incarceration <- round(sd(cleaned_data$incarceration_rate, na.rm = TRUE), 2)
median_incarceration <- round(median(cleaned_data$incarceration_rate, na.rm = TRUE), 2)

mean_poverty <- round(mean(cleaned_data$poverty_rate, na.rm = TRUE), 2)
sd_poverty <- round(sd(cleaned_data$poverty_rate, na.rm = TRUE), 2)
median_poverty <- round(median(cleaned_data$poverty_rate, na.rm = TRUE), 2)

mean_unemployment <- round(mean(cleaned_data$unemployment_rate, na.rm = TRUE), 2)
sd_unemployment <- round(sd(cleaned_data$unemployment_rate, na.rm = TRUE), 2)
median_unemployment <- round(median(cleaned_data$unemployment_rate, na.rm = TRUE), 2)

mean_violent <- round(mean(cleaned_data$violent_crime_rate, na.rm = TRUE), 2)
sd_violent <- round(sd(cleaned_data$violent_crime_rate, na.rm = TRUE), 2)
median_violent <- round(median(cleaned_data$violent_crime_rate, na.rm = TRUE), 2)

# Create a data frame manually
data <- data.frame(
  Variable = c("Incarceration Rate", "Poverty Rate", "Unemployment Rate", "Violent Crime Rate"),
  Mean = c(mean_incarceration, mean_poverty, mean_unemployment, mean_violent),
  Median = c(median_incarceration, median_poverty, median_unemployment, median_violent),
  SD = c(sd_incarceration, sd_poverty, sd_unemployment, sd_violent)
)

# Print the data frame as a table using kable
kable(data, col.names = c("Variable", "Mean", "Median", "SD"), booktabs = TRUE)
```

As it stands, by adhering to the figures above, there does not appear to be a strong correlation between any one of the variables and incarceration rates. In particular, poverty rates and unemployment rates do not showcase any particular trend in any direction. Violent crime on the other hand, does appear to show a somewhat positive correlation with incarceration rates. However what is interesting to note is that the 5 highlighted nations spanning the entire range of incarceration rates do not span a significant range of violent crime rates, at least relative to the overall range of violent crime rates across the entire dataset. The educational ranking category appears to have a negative relationship with incarceration rates, where a higher rank, i.e. a better education, seems to correspond with a lower average incarceration rate. However, somewhat interesting to note is the drop in mean incarceration rates for categories past the peak in the "111-125" ranking category. As such, given the fairly inconclusive information provided by the graphical and tabular figures above regarding the relationship between the chosen socioeconomic factors and a nation's incarceration rate, we employ a model in @sec-model to come to a more conclusive result.

# Model {#sec-model}

As the purpose of this paper is the investigation of various socioeconomic factors and their impact on a nation's incarceration rate as a means of better understanding the United States' mass incarceration problem, we will employ a model that will better enable us to infer the relationship between these various factors and incarceration rates. Then, using this information, I will attempt to predict what every nation's, including the United States', incarceration rate should be given its own individual data for each of the determinants of incarceration rate, and subsequently uncover whether these socioeconomic factors are a good estimator of incarceration rate, or whether issues regarding mass incarceration stem from a different facet. 

## Model set-up

The particular model that this paper will utilize is a Bayesian multiple linear regression model of incarceration rate as a function of the socioeconomic variables present in our analysis. In particular, the model is as follows:

\begin{align} 
y_i|\mu_i, \sigma &\sim \mbox{Normal}(\mu_i, \sigma) \\
\mu_i &= \beta_0 + \beta_1 poverty_{i} + \beta_2 crime_{i} + \beta_3 unemployment_{i} + \gamma_1 educ_{i} + ... + \gamma_9 educ_{i} \\
\beta_0 &\sim \mbox{Normal}(0, 2.5) \\
\beta_1 &\sim \mbox{Normal}(0, 2.5) \\
\beta_2 &\sim \mbox{Normal}(0, 2.5) \\
\beta_3 &\sim \mbox{Normal}(0, 2.5) \\
\gamma_i &\sim \mbox{Normal}(0, 2.5) \\
\sigma &\sim \mbox{Exponential}(1)
\end{align}

In the above model:

- $\mu_i$ is the predicted incarceration rate for a nation given a nation's poverty rate, violent crime rate, unemployment rate, and educational ranking category.
- $\beta_0$ is the coefficient for the intercept.
- $\beta_1$ is the coefficient for the predicted change in the incarceration rate of a nation given a one unit, i.e. one percentage point, increase in the nation's poverty rate.
- $\beta_2$ is the coefficient for the predicted change in the incarceration rate of a nation given a one unit increase in the nation's violent crime rate.
- $\beta_3$ is the coefficient for the predicted change in the incarceration rate of a nation given a one unit, i.e. one percentage point, increase in the nation's unemployment rate.
- $\gamma_i$, for $i = 1, ..., 9$, represents the coefficient for the predicted change in the incarceration rate of a nation given that it is in one of the educational ranking categories, "20-34", ..., "176-198", respectively for each $i$. Notice, this implies that the base intercept assumes that a nation is in the "1-19" educational ranking category.

Note, we employ `rstanarm` [@citerstanarm] in order to run the regression. Moreover, we make use of the default priors offered by `rstanarm`, however we allow for auto scaling.

### Model justification

The above factors were each chosen due to the literature present regarding each of their impacts on incarceration. In particular, there is data to suggest that more impoverished individuals are more susceptible to being arrested, at least in the United States [@citePovertyImpact]. Moreover, there is research to suggest that unemployment rates seem to correlate with incarceration rates, at least over extended periods of time in the United States [@citeUnemploymentImpact]. There is also literature that suggests that education significantly reduces the likelihood of incarceration, with claims that incarceration rates monotonically decline with education [@citeEducationImpact]. As for the impact of violent crime rates on incarceration rates, most research generally tends to focus on the reverse, focusing on incarceration rates' impact on violent crime. As such, this paper seeks to explore the converse in order to gain a more well-rounded understanding of how both factors influence one another on the whole.

A priori, we expect a linear relationship between each of these variables and a nation's incarceration rate. More specifically, we would expect that as the the level of violent crime, poverty, and unemployment rises, so too does the incarceration rate, as individuals may find themselves more likely to commit crimes as a means of survival, in turn increasing the amount of individuals that become imprisoned. On the contrary, we expect that more educated nations will generally see lower incarceration rates as better education is likely to be correlated with better standards of living and lower crime rates. 

# Results {#sec-results}

## Model Coefficients {#sec-modelcoefficients}

After running the regression based on the above model, we receive the following coefficient values as showcased in @tbl-incarcerationsummary. 

```{r}
#| label: tbl-incarcerationsummary
#| tbl-cap: Model summary of the predicted impact of crime, education, unemployment, and poverty on incarceration rates.
#| echo: false
#| warning: false
#| message: false
#| tbl-pos: H

modelsummary(analysis_model)

```

We can see that, somewhat unexpectedly, increases in the poverty rate negatively correlate with incarceration rates, as do increases in unemployment, assuming all else held constant. On the contrary, as predicted, violent crime rates correlate positively with incarceration rates, all else held constant. As for a nation's educational ranking, it appears that there is no clear relationship between education and incarceration rates as nations in the highest category of education, "1-19", would see a much larger increase in predicted incarceration rates as they enter categories near the median educational ranking, before then seeing smaller increases near the lowest ranked nations.

## Actual vs. Predicted Incarceration Rates {#sec-actualvspredicted}

In order to get a better grasp of the predictive power of our model, we compare every nation's predicted incarceration rate, given their respective qualities, to their actual incarceration rate as showcased in @fig-actualvspredicted below.

```{r}
#| label: fig-actualvspredicted
#| tbl-cap: Comparison of actual and predicted incarceration rates.
#| echo: false
#| fig-pos: H

# Extract coefficient estimates
coefficients <- coef(analysis_model)

# Extract predictor variables (poverty, crime, unemployment, and education)
predictors <- cleaned_data |>
  select(poverty_rate, violent_crime_rate, unemployment_rate, education_category)

# Calculate the predicted incarceration rate for each nation
predicted_incarceration <- (
  coefficients["(Intercept)"] +                      
  coefficients["poverty_rate"] * predictors$poverty_rate +
  coefficients["violent_crime_rate"] * predictors$violent_crime_rate +
  coefficients["unemployment_rate"] * predictors$unemployment_rate +
  coefficients["education_category20-34"] * as.numeric(predictors$education_category == "20-34") +
  coefficients["education_category35-52"] * as.numeric(predictors$education_category == "35-52") +
  coefficients["education_category53-71"] * as.numeric(predictors$education_category == "53-71") +
  coefficients["education_category72-92"] * as.numeric(predictors$education_category == "72-92") +
  coefficients["education_category93-110"] * as.numeric(predictors$education_category == "93-110") +
  coefficients["education_category111-125"] * as.numeric(predictors$education_category == "111-125") +
  coefficients["education_category126-147"] * as.numeric(predictors$education_category == "126-147") +
  coefficients["education_category148-175"] * as.numeric(predictors$education_category == "148-175") +
  coefficients["education_category176-198"] * as.numeric(predictors$education_category == "176-198") 
)

# Add predicted incarceration rate as a new column in cleaned_data
cleaned_data <- cleaned_data |>
  mutate(predicted_incarceration_rate = predicted_incarceration)

# Graph the resultant scatter plot
ggplot(cleaned_data, aes(x = predicted_incarceration_rate, y = incarceration_rate)) +
  geom_point(color = "darkgrey") +  # Add points for actual incarceration rates
  geom_abline(intercept = 0, slope = 1, color = "black", linetype = "dashed") +  # Add y=x line
  geom_point(data = cleaned_data |> filter(country %in% c("United States", "El Salvador", "Guinea-Bissau", "Iceland", "Mexico")),
             aes(color = country), size = 1.5) +
  labs(x = "Predicted Incarceration Rate", y = "Actual Incarceration Rate") + 
  theme_classic()  

```

In the above figure, each point represents a given nation's actual incarceration rate (on the y-axis), against their predicted incarceration rate (on the x-axis) based on our employed model. We include a dotted $y=x$ line for reference to showcase all the theoretical points where the actual incarceration rate of a nation equals the predicted incarceration rate. Note that any points below the dotted $y=x$ line signify nations whose predicted incarceration rates exceeds their actual incarceration rates, whereas points above the line signify nations whose actual exceeds their predicted. 

Examining the results on the whole, it appears that the model is able to somewhat accurately predict a nation's incarceration rate, whenever that nation's actual incarceration rate is between 100 and 300. As a nation's actual incarceration rises above 300, the model loses its predictive power as it appears to significantly underestimate the nation's incarceration rate. On the other hand, with nations whose actual incarceration rates are below 100, it appears that model generally tends to overestimate the nation's incarceration rate. 

We highlight the same 5 nations as done in @sec-data as they span the entire range of incarceration rates. We can see that, as observed, Iceland and Guinea-Bissau, who have the 5th lowest and lowest incarceration rates of 31 and 36, respectively, feature predicted incarceration rates that exceed their actual. El Salvador and the United States, who have the highest and 3rd highest incarceration rates of 1086 and 531, respectively, featured predicted incarceration rates far below their actual. On the other hand, Mexico, with an actual incarceration rate of 174, had a predicted incarceration rate almost exactly equal to its actual rate. Thus, these 5 nations effectively showcase where the model's predictive power is strongest - near the middle - and where it is weakest - on either extreme.

We discuss the implications of these results in @sec-discussion, and in particular, how this helps in understanding the United States' issue of mass incarceration.

# Discussion {#sec-discussion}

## Nations with extreme incarceration rates highlight the importance of cultural factors in determining their incarceration rates {#sec-culture}

As the results in @sec-actualvspredicted have shown, nations with excessively high and excessively low incarceration rates have various qualities about them, outside the scope of our model, that may better explain how and why they may find themselves with the incarceration rates that they do have. In particular, various cultural differences across nations such as differences in how nations view the idea of incarceration could significantly impact how various countries choose to deal with crime. 

Countries such as Norway prioritize rehabilitation over punitive measures, leading to recidivism rates as low as 20%. On the contrary, the United States - who focus much more heavily on punishment than rehabilitation - finds over two-thirds of released prisoners returning back to the criminal justice system within 2 years [@citeNorwayUS]. Such stark differences in how nations choose to deal with crime can vastly impact how likely they are to incarcerate an individual, as well as how likely incarcerated individuals are to return to crime. In essence, prioritizing punishment as a deterrent can create a vicious cycle of high incarceration rates subsequently leading to high crime rates, in turn necessitating the need for more punishment, continuing the cycle.

Thus, it could appear that countries who suffer from mass incarceration, such as the United States, may find themselves in these positions by their own volition, as they choose to focus on harming their prisoners than helping.

## Economic hardships and corruption at the national level may prevent nations from incarcerating when necessary {#sec-povertycorruption}

The results in @tbl-incarcerationsummary indicated a negative relationship between poverty rates and incarceration rates, suggesting that nations with a generally poorer population, relative to that particular nation's cost of living, and higher income inequality may find themselves with lower incarceration rates. Though this may seem counter-intuitive at first glance, a potential explanation could be that poorer nations are simply less capable of handling crime. In particular, poorer nations may not have the funding necessary to house criminals or hire a sufficient number of individuals to police the nation. For instance Kenya, who feature a poverty rate of 36.1% - well above the median - but an incarceration rate of 107 - fairly below the median, also happen to have a jail occupancy level of 173.2% as of 2022 [@citeKenyaJail]. This could help explain Kenya's relatively low incarceration rate as being the result of an inability to house more inmates.

Another important point to raise is the effect of national poverty on the extent of corruption within a nation. Studies generally show that the levels of corruption in developing nations outweigh those of developed nations [@citeCorruption]. As a result, it may be the case that poorer nations find themselves with lower incarceration rates as individuals are more easily capable of avoiding imprisonment through means such as bribery. In turn, extreme poverty can in turn lead to a lack of incarceration both due to financial concerns as well as higher rates of corruption.

## Educational attainment as a nuanced metric of incarceration rates {#sec-educationincarceration}

As was explored in @sec-modelcoefficients, we found that there was no consistent relationship between educational ranking and incarceration rates. In particular, though the highest educationally ranked nations found themselves with lower incarceration rates than those with poorer education, the extent to which incarceration rates were predicted to rise with a decrease in education varied depending on how much the educational ranking would fall. Somewhat counter intuitively, a nation falling from the highest educational category to that of the ranks "111-125" would find themselves with the highest increase in incarceration rates compared to a further worsening of education to even the lowest category of "176-198". 

The possible explanations for this are twofold: Firstly, as was previously examined in @sec-culture, incarceration rates can be heavily dependent on cultural aspects that were not taken into account via the model. As such, certain nations with exceedingly low or exceedingly high incarceration rates may be disproportionately affecting the mean incarceration rates of their respective educational ranking categories, in turn leading to inaccurate conclusions regarding the impact of education on incarceration rates. To illustrate this point further, El Salvador, who feature the highest incarceration rate of 1086, also happen to be in the educational ranking category of "111-125", which may explain why the model predicts that this particular category is likely to raise a nation's incarceration rate by the greatest margin. 

The second possible explanation could be that nations with poorer levels of education are likely to also be poorer economically, which, as discussed in @sec-povertycorruption, could lead to lower incarceration rates, in turn providing a possible explanation for why nations with the lowest levels of education also happen to feature lower incarceration rates than some nations with better education. 

## Insights for addressing the mass incarceration challenge in the United States drawn from the data and results

By incorporating all insights derived from @sec-culture, @sec-povertycorruption, and @sec-educationincarceration, we arrive to the following conclusions for how the United States can best address their issue of mass incarceration:

(1) The United States should stray from punitive measures as the primary means of dealing with crime as this can perpetuate a cycle of high incarceration rates, recidivism, and crime. Instead, shifting towards rehabilitation-oriented policies, such as those endorsed by nations like Norway, can be more effective in reducing the need for incarceration, and reduce the likelihood of repeat-offenders.

(2) The United States should dedicate more resources towards enhancing their educational prowess. As it stands, the United States has an education ranked between "20-34", and as a result, given the overwhelming evidence that nations with the highest educational strength also happen to have the lowest incarceration rates, an increase in investment in education can aid the United States in decreasing both their violent crime rate, as well as the need for mass incarceration. Though we have uncovered that education can be closely linked to other factors of a nation including economic strength and standards of living, the United States is one of the most economically powerful nations, and as a result, we may come to the justified conclusion that improvements in their education are likely to yield positive results akin to that of similar nations.


# Limitations and Next Steps {#sec-limitations}

## Limitations

There are a number of limitations to be addressed with regards to the analysis conducted and subsequent conclusions drawn. A significant limitation pertaining to this study is the lack of up-to-date data across all nations. As was discuss is @sec-data, all raw datasets contained information on nations that may not be accurate as of 2024. In particular, many nations had data that was collected years, and in some cases decades, ago, in turn potentially rendering much of the analysis inaccurate as significant changes could have risen over the years. Moreover, not all datasets included information on all nations, and in many cases, countries that were included in some datasets featured missing data points for various columns of data. As a result, the subset of countries that were included in this study may not accurately reflect the overarching global trends as only 134 out of the 193 countries recognized by the UN [@citeUN] were taken into account during the analysis. Thus, the subsequent conclusions drawn may be biased as the remaining countries left in the study could be skewed towards more developed nations as these may be more likely to have data readily available. Moreover, due to the missing data points for countries that were included in the study, values such as the violent crime rate statistic that was calculated for each nation is inaccurate as it only included whichever information was available for that particular nation. As a result, nations which may have reported more crime rates than others would in turn bias the results to inaccurately portray such nations as having relatively more crime than they should when compared with nations that had missing data points for certain crime statistics. 

A potentially significant limitation also exists with regards to the model used. In particular, the model does not take into account any interaction terms between the various independent variables when it likely should have, given the complex and intricate dynamics between the many variables at play. Though this was done to simplify the complexity of the model to allow for easier interpretations of the results, the caveat was an overall less realistic, and possibly less accurate, model. Another limitation with the model lies in its omission of cultural variations and associated factors across nations. By not grouping countries with similar cultural values or political and judicial policies, the model fails to account for factors that could greatly influence the derived results. As we have already explored in @sec-culture, varying cultural and societal norms with regards to how nations deal with crime can significantly impact their incarceration rates - a fact not taken into account by the model.

## Next Steps

Future analysis in the context of this study can be better improved by incorporating more up-to-date data. In particular, web-scraping and compiling the raw data manually as opposed to using the raw datasets used in this study could net an overall more recent, and complete, set of data points. Moreover, incorporating time series data spanning multiple years across all nations would allow us to account for time and country fixed-effects, in turn providing more conclusive data that takes into account the various cultural and societal disparities across nations and years. Another improvement that could be made is a better analysis of the model itself in order to fine-tune aspects such as the locations of the various priors, as well as exploring the possibility of adding additional factors and interaction terms to more accurately model incarceration rates.

\newpage

\appendix

# Appendix {-}

# Model details {#sec-model-details}

## Posterior Predictive Check

We employ a posterior predictive check in @fig-ppcheck to evaluate how well the model fits the data. By examining the figure, it appears as though the model does not accurately fit the data, and thus amendments likely should be made as a result [@citeAlexander].

```{r}
#| label: fig-ppcheck
#| fig-cap: Examining how the model fits the data through a posterior predictive check
#| echo: false
#| fig-pos: H
#| message: false
#| warning: false

pp_check(analysis_model) + 
  theme_classic() +
  theme(legend.position = "bottom")
```

## Comparison of the Posterior and Prior

We compare the posterior and prior in @fig-postprior to examine how the estimates change once data is taken into account [@citeAlexander].

```{r}
#| label: fig-postprior
#| fig-cap: Examining how the model is affected by the data through a posterior-prior comparison
#| echo: false
#| warning: false
#| fig-pos: H
#| message: false


posterior_vs_prior(analysis_model) + 
  theme_classic() + 
  theme(legend.position = "bottom") +
  coord_flip()
```

## Markov Chain Monte Carlo Convergence

As `rstanarm` makes use of the sampling algorithm MCMC [@citeAlexander], we examine whether the algorithm ran into issues through the use of an Rhat plot and trace plot, both showcased in @fig-mcmc. The Rhat plot does not indicate any issues, evidenced by the values all remaining close to 1 [@citeAlexander]. Moreover, the trace plot suggests nothing out of the ordinary given that lines remain horizontal and fluctuate in an expected manner [@citeAlexander].

```{r}
#| label: fig-mcmc
#| fig-cap: "Checking the convergence of the MCMC algorithm"
#| fig-subcap: ["Rhat plot", "Trace plot"]
#| echo: false
#| message: false
#| warning: false
#| fig-pos: H
#| layout-ncol: 1

plot(analysis_model, "rhat") + 
  theme_classic()

plot(analysis_model, "trace") 

```

## Credibility Intervals

Lastly, we employ a 95% credibility interval as showcased in @fig-ci, to gain a better understanding of the probability distribution of our coefficients.

```{r}
#| label: fig-ci
#| fig-cap: "Credible intervals"
#| echo: false
#| message: false
#| warning: false
#| fig-pos: H

plot(analysis_model) + 
  theme_classic()

```

\newpage


# References


